Reconstructing the neutron-star equation of state with gravitational-wave detectors from a realistic population of inspiralling binary neutron stars

### 2014-09-13 paper.tex
Rewrote Section II.

%Observations of neutron stars (NSs), consisting of nuclear matter in the ground state from zero to several times nuclear saturation density ($\rho_{\rm nuc}\sim2.8\times 10^{14}$~g/cm$^3$), in principle provide an ideal way to measure the nuclear equation of state (EOS) that describes the pressure $p$ as a function of density $\rho$. The most rigorous EOS constraints from NSs have come from measurements of high mass pulsars in compact binary systems, and it is now clear that the EOS must allow for NS masses $\gtrsim 2 M_\odot$~\cite{DemorestPennucciRansom2010, AntoniadisFreireWex2013}. More precise information can be obtained if the NS mass $M$ and radius $R$ are simultaneously measured, and Lindblom demonstrated explicitly that there is a one-to-one map between the relations $R(M)$ and $p(\rho)$~\cite{Lindblom1992}. However, different observations and analyses have produced only marginally consistent results. For example, \"Ozel~{\it et al.}\ used observations of thermonuclear bursts from x-ray binaries to measure the NS mass and radius in three systems. For the three systems considered, the 95\% confidence regions spanned the range $7.5$--$12$~km (Fig.~1 of Ref.~\cite{OzelBaymGuver2010}). Steiner~{\it et al.}\ also used mass and radius measurements from three [[two were the same as Ozel]] x-ray bursters as well measurements from three quiescent low-mass x-ray binaries and found the NS radius to be $10.7$--12.5~km (from the range of 95\% confidence intervals in Tables~7 and~8 of Ref~\cite{SteinerLattimerBrown2010}). Guillot~{\it et al.}\ measured the mass and radius for five quiescent low-mass x-ray binaries [[two were the same as Steiner et al.]], and after stacking the results found that the NS radius was 7.6--10.4km (90\% confidence) assuming the mass was an approximately constant function of mass~\cite{GuillotServillatWebb2013}.

%Observations of neutron stars (NSs), consisting of nuclear matter in the ground state from zero to several times nuclear saturation density ($\rho_{\rm nuc}\sim2.8\times 10^{14}$~g/cm$^3$), in principle provide an ideal way to measure the nuclear equation of state (EOS) that describes the pressure $p$ as a function of density $\rho$. The most rigorous EOS constraints from NSs have come from measurements of high mass pulsars in compact binary systems, and it is now clear that the EOS must allow for NS masses $\gtrsim 2 M_\odot$~\cite{DemorestPennucciRansom2010, AntoniadisFreireWex2013}. More precise information can be obtained if the NS mass $M$ and radius $R$ are simultaneously measured, and Lindblom demonstrated explicitly that there is a one-to-one map between the relations $R(M)$ and $p(\rho)$~\cite{Lindblom1992}. However, there has been disagreement about the reliability of NS radius measurements. Recent works using multiple NS observations have concluded that the radius of a canonical $1.4M_\odot$ NS, for example, is $9$--$12$~km (for the range of 95\% confidence intervals for three NSs shown in Fig.~1 of Ref.~\cite{OzelBaymGuver2010}), $10.7$--12.5~km (95\% confidence)~\cite{SteinerLattimerBrown2010}, and 7.6--10.4km (90\% confidence)~\cite{GuillotServillatWebb2013}. \les{This sentence confused me.  Is the ``multiple NS observations'' referring to three of the NSs from Fig 1, like you state in the parenthesis?  Is the point that these three papers independently used observations from three of the NSs in Fig 1 (did they use the same three?) to estimate the radius of a $1.4M_\odot$ NS, and all came up with different ranges?}{}

@@

The measurability of tidal parameters in BNS systems with aLIGO for GW frequencies below 450Hz (prior to the last $\sim 20$ GW cycles before merger where higher order corrections to the tidal effect become important) was first examined for polytropic EOSs in~\cite{FlanaganHinderer2008} and for theoretical hadronic and quark matter EOSs in~\cite{HindererLackeyLangRead2010}.  These studies found that for a single aLIGO detector, tidal interactions were only observable during this early inspiral stage for stiff EOSs, NS masses below 1.4~$M_\odot$, and for \ben{rare}{}, nearby sources with signal-to-noise ratios (SNR) $\gtrsim 30$. Damour {\it et al.}, however, found that when also including EOS information from the last $\sim 20$ GW cycles prior to contact, tidal parameters are in fact observable \ben{even with more common}{ \les{for more likely}{}} SNRs of $\sim 16$~\cite{DamourNagarVillain2012}. 
%\les{The highlighted phrase in the previous sentence confused me.  Are you saying that, by including information from the last 20 cycles, the signals are now observable at SNR~16, and that SNR~16 signals are much more likely than SNR~30 signals? }{}

@@

%In this paper we will demonstrate the advantages of parameterizing the EOS instead of $\lambda(M)$. The main benefit comes from including prior information about the EOS parameters that would be more difficult to include for the parameters of a $\lambda(M)$ fit. As will be discussed in Section~\ref{sec:eos}, thermodynamic stability and causality provide the simple constraint that the slope of the EOS must be in the range $0 \le dp/d\epsilon \le c^2$, where $\epsilon$ is the energy density \les{and $c$ is the speed of light}{}. Additionally, the constraints from the observed $2M_\odot$ NSs can be included by simply rejecting any EOS parameters that lead to a NS maximum mass below $2M_\odot$. We will find that incorporating this information allows us to make significantly stronger statements about the EOS, radius, and tidal deformability, and this is true for a narrower, more realistic range of NS masses than simulated by Del Pozzo {\it et al.}~\cite{DelPozzoLiAgathos2013}.

The BNS inspiral signal is almost exactly characterized by the mass, spin, and tidal deformability of each NS, and is therefore free from the intrinsic variability and uncertainties that bias electromagnetic measurements of the mass and radius. Currently unknown terms in the post-Newtonian waveform model, however, will lead to systematic errors in recovering the tidal deformabilities \ben{}{and, to a lesser extent, masses}. Using the Fisher matrix approximation~\cite{Favata2014, YagiYunes2014} as well as a Markov chain Monte Carlo (MCMC) Bayesian analysis~\cite{WadeCreightonOchsner2014}, it was found that the systematic error in $\lambda$ from waveform uncertainties can be as large as the statistical error for a single BNS inspiral observation. When stacking observations the problem becomes worse, as the statistical errors decrease with more observations but the systematic errors do not, and we will discuss how systematic errors impact the recovery of the EOS.

@@

\red{[[Make sure you cite Damour's papers (two for 4PN energy, and the TaylorXy convention). See his email.]]}

\les{Some citations still needed...  (Ben, you mention an email from Damour.  Do you want to add in those citations?)}{}

\les{Check that notation is consistent with rest of paper.}{}

\les{Should I include $c$'s and $G$'s?}{}

The strain $h$ observed in a GW detector from a wave emitted by a BNS system is related to the two polarizations of the wave ($h_+$ and $h_\times$) and to the detector's antenna beam pattern response ($F_+$ and $F_\times$).  Assuming the binary components are point particles in a circular orbit and restricting to leading order in amplitude, the two polarizations of the emitted GW seen by an observer at a distance $D$ away are
\begin{eqnarray}
\label{eq:hoftp}
h_+(t)&=&-\frac{4\mu}{D}\left(\frac{1+\cos^2\iota}{2}\right)x(t)\cos\left[2\Phi(t)\right],\\
\label{eq:hoftc}
h_\times(t)&=&-\frac{4\mu}{D}\cos\iota\mbox{ }x(t)\sin\left[2\Phi(t)\right],
\end{eqnarray}
in the time-domain, where $\mu=m_1 m_2/M$ is the reduced mass, $M=m_1+m_2$ is the total mass, $\iota$ is the inclination angle \les{Should I define what the inclination angle is?}{}, $\Phi(t)$ is the orbital phase, and $x=(\pi M f)^{2/3}$ where $f$ is the GW frequency.  Using the Fourier transform convention
\begin{equation}
\tilde{h}(f)=\int_{-\infty}^{+\infty} h(t) e^{-2\pi i f t}dt,
\end{equation}
the stationary-phase approximation can be used to find the two polarizations in the frequency-domain:
\begin{eqnarray}
\label{eq:hoffp}
\tilde{h}_+(f)&=&\sqrt{\frac{5}{24}}\frac{{\cal M}^{5/6}f^{-7/6}}{\pi^{2/3}D}\left(\frac{1+\cos^2\iota}{2}\right)e^{-i\Psi(f)},\\
\label{eq:hoffc}
\tilde{h}_\times(f)&=&-i\sqrt{\frac{5}{24}}\frac{{\cal M}^{5/6}f^{-7/6}}{\pi^{2/3}D}\cos\iota\mbox{ }e^{-i\Psi(f)},
\end{eqnarray}
where the chirp mass ${\cal M}=\eta^{3/5}M$, the symmetric mass ratio $\eta=m_1 m_2/M^2$, $\Psi(f)=2\pi f t(v) - 2\Phi(v) - \pi/4$, and $\tilde{h}_+(f)$ and $\tilde{h}_\times(f)$ are the Fourier transforms of $h_+(t)$ and $h_\times(t)$ respectively.  Given an energy $E$ and luminosity ${\cal L}$, $\Phi(t)$ and $x(t)$, which determines the two time-domain GW polarizations in Eqs.~\eqref{eq:hoftp} and~\eqref{eq:hoftc}, and $\Psi(f)$, which determines the two frequency-domain GW polarizations in Eqs.~\eqref{eq:hoffp} and~\eqref{eq:hoffc}, can be found from the following ordinary differential equations:
\begin{eqnarray}
\label{eq:dPhidt}
\frac{d\Phi(t)}{dt}&=&\frac{x^{3/2}}{M},\\
\label{eq:dxdt}
\frac{d x(t)}{dt}&=&\frac{-{\cal L}}{E^{\prime}},
\end{eqnarray}
where $E^{\prime}=dE/dx$ is the analytic derivative of the energy $E$, and the luminosity ${\cal L}$ is related to energy via the energy balance requirement $dE/dt=-{\cal L}$.

The energy and luminosity of two inspiralling point-particles can be easily solved for in the Newtonian limit of General Relativity.  However, as the system approaches merger, the orbital speed becomes relativistic, and this quasi-Newtonian limit is not enough to accurately describe its waveform.  We therefore use post-Newtonian (PN) theory, which involves expansions of the equations of motion in orders of orbital speed, to push beyond this limit.  Furthermore, compact binaries containing a neutron star noticeably depart from the point-particle approximation at high frequencies and require an additional correction to the energy and luminosity of the system relative to the point-particle terms.  More specifically, tidal interactions contribute to the phase of the orbit by adding a monopole-quadrupole interaction term to the energy.  This additional quadrupole moment adds to the total quadrupole moment of the system, which also increases the luminosity.

As was done in most previous works~\cite{WadeCreightonOchsner2014, DelPozzoLiAgathos2013, Favata2014, YagiYunes2014}, we keep the PN point-particle corrections to the energy and luminosity up to 3.5PN order as well as the leading-order and next-to-leading-order tidal contributions, which enter at the same order as 5PN and 6PN point-particle terms respectively.  The energy and luminosity take the form:
\begin{align}
E &= - \frac{1}{2} M \eta x \left[1 + e_\text{PP-PN}(\eta) + e_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right], \\
\mathcal{L} &= \frac{32}{5}\eta^2 x^5 \left[ 1 + l_\text{PP-PN}(\eta) + l_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right],
\end{align}
where the point-particle PN terms $e_\text{PP-PN}$ and $l_\text{PP-PN}$ are given in Ref.~\cite{Blanchet2014Review}, and the tidal terms are given in Ref.~\cite{WadeCreightonOchsner2014}.  In order to write $e_\text{Tidal}$ and $l_\text{Tidal}$ as a function of the dimensionless quantity $\eta$, we have asserted that $m_1 \ge m_2$.

As is described in detail in the Appendix of Ref.~\cite{WadeCreightonOchsner2014}, there are several methods to determine the phase evolution of the system from its energy and luminosity, and we use three of them.  In the TaylorT1 method, one numerically integrates Eqs.~\eqref{eq:dPhidt} and~\eqref{eq:dxdt} to find $\Phi(t)$ and $x(t)$. The TaylorT4 method is a slight variation where one re-expands the ratio $-{\cal L}/E^\prime$ on right hand side of Eq.~\eqref{eq:dxdt} then truncates the series at the highest known PN order (here 3.5PN) before numerically integrating to find $\Phi(t)$ and $x(t)$. The TaylorF2 method uses the stationary phase approximation to estimate the Fourier transformed waveform.  Eqs.~\eqref{eq:dPhidt} and~\eqref{eq:dxdt} are analytically integrated to write $\Psi(f)$ in the following form:
\begin{equation}
\label{eq:psi}
\Psi(f)=2\pi f t_{\rm ref} - 2\Phi_{\rm ref} -\frac{\pi}{4}+ 2\int_x^{x_{\rm ref}}\frac{x^{3/2} - y^{3/2}}{M}\frac{E^\prime(y)}{{\cal L}(y)}dy,
\end{equation}
where $t_{\rm ref}=t(x_{\rm ref})$, $\Phi_{\rm ref}=\Phi(x_{\rm ref})$, and $x_{\rm ref}$ is an arbitrary reference value, following \cite{BuonannoIyerOchsner2009}.  The ratio $E^\prime/{\cal L}$ is then re-expanded and truncated at the highest known PN order (here 3.5PN) before the integral in Eq.~\eqref{eq:psi} is analytically computed to find $\Psi(f)$.  The differences in the PN waveform families are due mostly to the way terms beyond 3.5PN order are truncated.  Thus, the differences in results using one waveform family as the injection and another as templates should give a reasonable estimate of systematic errors due to unknown higher order terms.

The majority of this work uses the TaylorF2 waveform approximant because its fully analytic frequency-domain form makes it the fastest PN waveform to generate.  The form of the phase $\Psi(f)$ with point-particle terms up to 3.5PN order and leading-order and next-to-leading-order tidal contributions is
\begin{eqnarray}
\Psi(f)&=&2\pi f t_{\rm ref} - 2\Phi_{\rm ref} -\frac{\pi}{4}\\
\nonumber
&&+\frac{3}{128\eta x^{5/2}}\left[1+\psi_\text{PP-PN}(\eta)+\psi_\text{Tidal}(\eta,\Lambda_1,\Lambda_2)\right],
\end{eqnarray}
where the point-particle PN terms $\psi_\text{PP-PN}(\eta)$ are given in \cite{BuonannoIyerOchsner2009}, and
\begin{widetext}
\begin{equation}
\label{eq:psi_tidal}
\psi_\text{Tidal}(\eta,\Lambda_1,\Lambda_2)=\left[-\frac{39}{2}\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)\right]x^5+ \left[-\frac{3115}{64}\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)+\frac{6595}{364}\sqrt{1-4\eta}\mbox{ }\delta\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)\right]x^6,
\end{equation}
where $\Lambda_1=\lambda_1/m_1^5$, $\Lambda_2=\lambda_2/m_2^5$, and
\begin{eqnarray}
\label{eq:LT}
\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)&=&\frac{8}{13}\left[\left(1+7\eta-31\eta^2\right)\left(\Lambda_1+\Lambda_2\right)+\sqrt{1-4\eta}\left(1+9\eta-11\eta^2\right)\left(\Lambda_1-\Lambda_2\right)\right],\\
\label{eq:dLT}
\delta\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)&=&\frac{1}{2}\left[\sqrt{1-4\eta}\left(1-\frac{13272}{1319}\eta+\frac{8944}{1319}\eta^2\right)\left(\Lambda_1+\Lambda_2\right) \right .\\
&& + \left . \left(1-\frac{15910}{1319}\eta+\frac{32850}{1319}\eta^2+\frac{3380}{1319}\eta^3\right)\left(\Lambda_1-\Lambda_2\right)\right].
\end{eqnarray}
\end{widetext}
We express the tidal contributions to the TaylorF2 phase in terms of $(\tilde\Lambda,\delta\tilde\Lambda)$ because $\tilde\Lambda$ is comparably more measurable than $\Lambda_1$ and $\Lambda_2$ individually.  For more details regarding this parameterization, see \cite{WadeCreightonOchsner2014,Favata2014}.

\ben{For simplicity, we choose to examine nonspinning BNS systems and do not include the NS spins as waveform parameters in this study.  However, it has been shown that not including spin parameters can noticeably bias parameter estimation even for systems with relatively small spin magnitudes \cite{Favata2014}.  Additionally, we choose to end our inspiral waveforms when they reach the GW frequency for the Schwarzschild innermost stable circular orbit (ISCO) $f_{\rm ISCO}=(6^{3/2}\pi M)$.  However, for large NS radii corresponding to a stiff EOS, NSs can merge before reaching $f_{\rm ISCO}$.  A parallel investigation which studies how spin and the choice of high-frequency cutoffs can affect the estimation of tidal parameters is nearing completion \cite{Nikhef_paper_2}.}{}

\les{}{We choose to ignore spin and precession in this study for simplicity.  However, it has been shown that ignoring spin in BNS systems can noticeably bias parameter estimation, even for relatively small spin magnitudes \cite{Favata2014}.  Additionally, we choose to end our waveforms when they reach $f_{\rm ISCO}=(6^{3/2}\pi M)$, where $f_{\rm ISCO}$ is the GW frequency of the innermost stable circular orbit (ISCO) of a test particle about a Schwarzschild BH of mass $M$.  For certain EOSs, this termination frequency may be greater than the frequency at which the two components come into contact.  A parallel investigation which studies how spin and the choice of high-frequency cutoffs can affect the estimation of tidal parameters is nearing completion \cite{Nikhef_paper_2}.}

%In this paper we use the post-Newtonian (PN) formalism to construct our waveform model [[as does everyone else]]. In the PN formalism the energy for circular orbits is found from the conservative parts of the Hamiltonian which is now known up to 4PN order [[is this true? Define what xPN order means]] The luminosity is found from a multipolar decomposition of the metric perturbation in the wave zone, which is then matched to the post-Newtonian multipolar decomposition in the near zone \red{[[None of this really matters.]]}. The phase evolution of the orbit is calculated from the assumption of energy balance $dE/dt = -\mathcal{L}$.

%Tidal interactions contribute to the phase of the orbit by adding a monopole-quadrupole interaction term to the energy, and the additional quadrupole moment adds to the total quadrupole moment of the system, increasing the luminosity. As was done in most previous works~\cite{WadeCreightonOchsner2014, DelPozzoLiAgathos2013, Favata2014, YagiYunes2014} we will use the point-particle terms to 3.5PN order as well as the leading and next-to-leading tidal contributions which enter the energy and luminosity at the same order as 5 and 6PN point-particle terms respectively.
%
%The energy and luminosity have the form
%\begin{align}
%E &= - \frac{1}{2} M \eta x \left[1 + e_\text{PP-PN}(\eta) + e_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right], \\
%\mathcal{L} &= \frac{32}{5}\eta^2 x^5 \left[ 1 + l_\text{PP-PN}(\eta) + l_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right],
%\end{align}
%where the point-particle PN terms $e_\text{PP-PN}$ and $l_\text{PP-PN}$ are given in Ref.~\cite{Blanchet2014Review}, and using the definition $m_1 \ge m_2$, the tidal terms are
%\begin{widetext}
%\begin{align}
%\begin{split}
%e_\text{Tidal} &= -\frac{9\eta}{2} \left[ (1-3\eta)(\Lambda_1+\Lambda_2) + \sqrt{1-4\eta}(1-\eta)(\Lambda_1-\Lambda_2) \right]x^5 \\
%& - 22\eta \left[ \left(1-4\eta+\frac{19\eta^2}{8}\right)(\Lambda_1+\Lambda_2) + \sqrt{1-4\eta}\left(1-2\eta+\frac{3\eta^2}{8}\right)(\Lambda_1-\Lambda_2) \right]x^6,
%\end{split} \\
%\begin{split}

>>>>

@@

The measurability of tidal parameters in BNS systems with aLIGO for GW frequencies below 450Hz (prior to the last $\sim 20$ GW cycles before merger where higher order corrections to the tidal effect become important) was first examined for polytropic EOSs in~\cite{FlanaganHinderer2008} and for theoretical hadronic and quark matter EOSs in~\cite{HindererLackeyLangRead2010}.  These studies found that for a single aLIGO detector, tidal interactions were only observable during this early inspiral stage for stiff EOSs, NS masses below 1.4~$M_\odot$, and for \ben{rare}{}, nearby sources with signal-to-noise ratios (SNR) $\gtrsim 30$. Damour {\it et al.}, however, found that when also including EOS information from the last $\sim 20$ GW cycles prior to contact, tidal parameters are in fact observable \ben{even with more common}{ \les{for more likely}{}} SNRs of $\sim 16$~\cite{DamourNagarVillain2012}.

@@

The BNS inspiral signal is almost exactly characterized by the mass, spin, and tidal deformability of each NS, and is therefore free from the intrinsic variability and uncertainties that bias electromagnetic measurements of the mass and radius. Currently unknown terms in the post-Newtonian (PN) waveform model, however, will lead to systematic errors in recovering the tidal deformabilities \ben{}{and, to a lesser extent, masses}. Using the Fisher matrix approximation~\cite{Favata2014, YagiYunes2014} as well as a Markov chain Monte Carlo (MCMC) Bayesian analysis~\cite{WadeCreightonOchsner2014}, it was found that the systematic error in $\lambda$ from waveform uncertainties can be as large as the statistical error for a single BNS inspiral observation. When stacking observations the problem becomes worse, as the statistical errors decrease with more observations but the systematic errors do not, and we will discuss how systematic errors impact the recovery of the EOS.

@@

The strain $h$ observed in a detector from a GW is related to the two polarizations of the GW, $h_+$ and $h_\times$, by the detector's antenna beam pattern response, $F_+$ and $F_\times$, through the relation
\begin{equation}
h = F_+ h_+ + F_\times h_\times.
\end{equation}
$F_+$ and $F_\times$ depend on the sky position (given by the right ascension $\alpha$ and declination $\delta$) and the GW polarization angle $\psi$.

For a BNS system in a quasicircular inspiral with an inclination angle $\iota$, luminosity distance $d_L$, and component masses $m_1$ and $m_2$, the two polarizations of the GW are
\begin{eqnarray}
\label{eq:hoftp}
h_+(t)&=&-\frac{4G\eta M}{c^2 d_L}\left(\frac{1+\cos^2\iota}{2}\right)x(t)\cos\left[2\Phi(t)\right],\\
\label{eq:hoftc}
h_\times(t)&=&-\frac{4G\eta M}{c^2 d_L}\cos\iota\mbox{ }x(t)\sin\left[2\Phi(t)\right],
\end{eqnarray}
where we have restricted the amplitude to the leading order and only keep the leading harmonic. Here, $M=m_1+m_2$ is the total mass, $\eta=m_1 m_2/M^2$ is the symmetric mass ratio, $\Phi$ is the orbital phase, and $x$ is the standard post-Newtonian parameter defined by $x = \left( \frac{GM}{c^3}\frac{d\Phi}{dt} \right)^{2/3} = \left( \frac{\pi GM f}{c^3} \right)^{2/3}$, where $f$ is the GW frequency.

The time evolution of $\Phi(t)$ and $x(t)$ are evaluated from the PN expressions for the energy $E$ and luminosity ${\cal L}$ via the energy balance requirement $dE/dt=-{\cal L}$:
\begin{eqnarray}
\label{eq:dPhidt}
\frac{d\Phi(t)}{dt}&=&\frac{c^3 x^{3/2}}{G M},\\
\label{eq:dxdt}
\frac{d x(t)}{dt}&=&\frac{-\mathcal{L}}{dE/dx}.
\end{eqnarray}
The energy and luminosity have the schematic form
\begin{align}
E &= - \frac{1}{2} c^2 M \eta x \left[1 + e_\text{PP-PN}(\eta) + e_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right], \\
\mathcal{L} &= \frac{32}{5} \frac{c^5}{G} \eta^2 x^5 \left[ 1 + l_\text{PP-PN}(\eta) + l_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right],
\end{align}
where we write the tidal terms as functions of the dimensionless quantities $\Lambda_i = G\lambda_i \left(\frac{c^2}{G m_i}\right)^5$

The point-particle PN terms for nonspinning systems depend on only $\eta$. The energy $e_\text{PP-PN}$ was recently calculated to 4PN order~\cite{BiniDamour2013, FoffaSturani2014}, and the luminosity $l_\text{PP-PN}$ is known to 3.5PN order~\cite{Blanchet2014Review}. For consistency with previous works~\cite{WadeCreightonOchsner2014, DelPozzoLiAgathos2013, Favata2014, YagiYunes2014}, we keep the point-particle PN corrections to 3.5PN order.

The leading tidal terms begin at the same order as 5PN point-particle terms, and although the tidal potential is known to the next-to-next-to-leading-order term~\cite{BiniDamourFaye2012}, we only include the leading and next-to-leading order terms for consistency with previous works~\cite{WadeCreightonOchsner2014, DelPozzoLiAgathos2013, Favata2014, YagiYunes2014}. These terms are~\cite{VinesFlanaganHinderer2011}
\begin{widetext}
\begin{align}
\begin{split}
e_\text{Tidal} &= -\frac{9\eta}{2} \left[ (1-3\eta)(\Lambda_1+\Lambda_2) + \sqrt{1-4\eta}(1-\eta)(\Lambda_1-\Lambda_2) \right]x^5 \\
& - 22\eta \left[ \left(1-4\eta+\frac{19\eta^2}{8}\right)(\Lambda_1+\Lambda_2) + \sqrt{1-4\eta}\left(1-2\eta+\frac{3\eta^2}{8}\right)(\Lambda_1-\Lambda_2) \right]x^6,
\end{split} \\
\begin{split}
l_\text{Tidal} &= 3 \left[ (1-2\eta-4\eta^2)(\Lambda_1+\Lambda_2) + \sqrt{1-4\eta}(1-2\eta^2)(\Lambda_1-\Lambda_2) \right]x^5 \\
& - \frac{22}{7} \left[ \left(1-\frac{15\eta}{176}+\frac{147\eta^2}{88}-\frac{1547\eta^3}{44}\right)(\Lambda_1+\Lambda_2) + \sqrt{1-4\eta}\left(1+\frac{337\eta}{176}+\frac{15\eta^2}{2}-\frac{1085\eta^3}{88}\right)(\Lambda_1-\Lambda_2) \right]x^6,
\end{split}
\end{align}
\end{widetext}
where $m_1 \ge m_2$.

As described in detail in Ref.~\cite{WadeCreightonOchsner2014}, there are several methods, first cataloged in Ref.~\cite{DamourIyerSathyaprakash2001}, to determine the phase evolution of the binary from the above expressions for its energy and luminosity, and we will use three of them.  In the TaylorT1 method, one numerically integrates Eqs.~\eqref{eq:dPhidt} and~\eqref{eq:dxdt} to find $\Phi(t)$ and $x(t)$ up to the time and phase constants $t_c$ and $\phi_c$. The TaylorT4 method is a slight variation where one re-expands the ratio $-{\cal L}/E^\prime$ in Eq.~\eqref{eq:dxdt} then truncates the series at the highest known PN order (here 3.5PN) before numerically integrating. Finally in the TaylorF2 method, one approximates the Fourier transform of the waveform
\begin{equation}
\tilde{h}(f)=\int_{-\infty}^{+\infty} h(t) e^{-2\pi i f t}dt
\end{equation}
with the stationary phase approximation. The result, to leading order in the amplitude, is
\begin{align}
\label{eq:hoffp}
\tilde{h}_+(f) &= \sqrt{\frac{5}{24}}\frac{G^2}{c^5}\frac{{\cal M}^{5/6}(Gf/c^3)^{-7/6}}{\pi^{2/3}d_L}\left(\frac{1+\cos^2\iota}{2}\right)e^{-i\Psi(f)},\\
\label{eq:hoffc}
\tilde{h}_\times(f) &= -i\sqrt{\frac{5}{24}}\frac{G^2}{c^5}\frac{{\cal M}^{5/6}(Gf/c^3)^{-7/6}}{\pi^{2/3}d_L}\cos\iota\mbox{ }e^{-i\Psi(f)}.
\end{align}
Here, $\mathcal{M} = (m_1 m_2)^{3/5}/M^{1/5}$ is the chirp mass and, for the TaylorF2 approach, the phase has the analytic form 
\begin{equation}
\begin{split}
\Psi(f) &= 2\pi f t_c - 2\phi_c -\frac{\pi}{4}\\
& +\frac{3}{128\eta x^{5/2}}\left[1+\psi_\text{PP-PN}(\eta)+\psi_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right],
\end{split}
\end{equation}
where the point-particle terms $\psi_{\rm PP-PN}$ are provided in~\cite{BuonannoIyerOchsner2009}. As was demonstrated in Refs.~\cite{WadeCreightonOchsner2014,Favata2014}, the individual tidal parameters $\Lambda_1$ and $\Lambda_2$ are highly correlated, so it is instead easier to reparameterize the tidal contribution to the phase $\psi_\text{Tidal}$ in terms of the linear combinations
\begin{align}
\begin{split}
\label{eq:LT}
\tilde{\Lambda}&=\frac{8}{13}\left[\left(1+7\eta-31\eta^2\right)\left(\Lambda_1+\Lambda_2\right)\right.\\
&+ \left.\sqrt{1-4\eta}\left(1+9\eta-11\eta^2\right)\left(\Lambda_1-\Lambda_2\right)\right],
\end{split}\\
\begin{split}
\label{eq:dLT}
\delta\tilde{\Lambda}&=\frac{1}{2}\left[\sqrt{1-4\eta}\left(1-\frac{13272}{1319}\eta+\frac{8944}{1319}\eta^2\right)\left(\Lambda_1+\Lambda_2\right) \right .\\
& + \left . \left(1-\frac{15910}{1319}\eta+\frac{32850}{1319}\eta^2+\frac{3380}{1319}\eta^3\right)\left(\Lambda_1-\Lambda_2\right)\right].
\end{split}
\end{align}
The tidal contribution then takes the simple form
\begin{equation}
\label{eq:psi_tidal}
\psi_\text{Tidal} = -\frac{39}{2}\tilde{\Lambda} x^5+ \left(-\frac{3115}{64}\tilde{\Lambda}+\frac{6595}{364}\sqrt{1-4\eta}\mbox{ }\delta\tilde{\Lambda}\right) x^6.
\end{equation}
The two terms containing $\tilde\Lambda$ are significantly larger than the term containing $\delta\tilde\Lambda$, and, as was previously found, $\delta\tilde\Lambda$ is not measurable with aLIGO~\cite{WadeCreightonOchsner2014}. This waveform model can therefore be expressed in terms of the 11 parameters $\vec\theta = \{d_L, \alpha, \delta, \psi, \iota, t_c, \phi_c, m_1, m_2, \tilde\Lambda, \delta\tilde\Lambda\}$.

For all versions of the PN waveform, we cut off the inspiral at the GW frequency corresponding to the Schwarzschild innermost stable circular orbit (ISCO) $f_{\rm ISCO}=(6^{3/2}\pi G M/c^3)$.  However, for large NS radii corresponding to a stiff EOS, NSs can merge before reaching $f_{\rm ISCO}$. For simplicity, we also choose to examine nonspinning BNS systems, although it has been shown that not including spin parameters can noticeably bias parameter estimation even for systems with relatively small spin magnitudes \cite{Favata2014}. A parallel investigation which studies how the choice of high-frequency cutoffs and spin can affect the estimation of tidal parameters is nearing completion \cite{Nikhef_paper_2} \red{[[Get reference]]}.

% ******************** start of Les' version ************************

%The strain $h$ observed in a GW detector from a wave emitted by a BNS system is related to the two polarizations of the wave ($h_+$ and $h_\times$) and to the detector's antenna beam pattern response ($F_+$ and $F_\times$).  Assuming the binary components are point particles in a circular orbit and restricting to leading order in amplitude, the two polarizations of the emitted GW seen by an observer at a distance $D$ away are
%\begin{eqnarray}
%\label{eq:hoftp}
%h_+(t)&=&-\frac{4\mu}{D}\left(\frac{1+\cos^2\iota}{2}\right)x(t)\cos\left[2\Phi(t)\right],\\
%\label{eq:hoftc}
%h_\times(t)&=&-\frac{4\mu}{D}\cos\iota\mbox{ }x(t)\sin\left[2\Phi(t)\right],
%\end{eqnarray}
%in the time-domain, where $\mu=m_1 m_2/M$ is the reduced mass, $M=m_1+m_2$ is the total mass, $\iota$ is the inclination angle \les{Should I define what the inclination angle is?}{}, $\Phi(t)$ is the orbital phase, and $x=(\pi M f)^{2/3}$ where $f$ is the GW frequency.  Using the Fourier transform convention
%\begin{equation}
%\tilde{h}(f)=\int_{-\infty}^{+\infty} h(t) e^{-2\pi i f t}dt,
%\end{equation}
%the stationary-phase approximation can be used to find the two polarizations in the frequency-domain:
%\begin{eqnarray}
%\label{eq:hoffp}
%\tilde{h}_+(f)&=&\sqrt{\frac{5}{24}}\frac{{\cal M}^{5/6}f^{-7/6}}{\pi^{2/3}D}\left(\frac{1+\cos^2\iota}{2}\right)e^{-i\Psi(f)},\\
%\label{eq:hoffc}
%\tilde{h}_\times(f)&=&-i\sqrt{\frac{5}{24}}\frac{{\cal M}^{5/6}f^{-7/6}}{\pi^{2/3}D}\cos\iota\mbox{ }e^{-i\Psi(f)},
%\end{eqnarray}
%where the chirp mass ${\cal M}=\eta^{3/5}M$, the symmetric mass ratio $\eta=m_1 m_2/M^2$, $\Psi(f)=2\pi f t(v) - 2\Phi(v) - \pi/4$, and $\tilde{h}_+(f)$ and $\tilde{h}_\times(f)$ are the Fourier transforms of $h_+(t)$ and $h_\times(t)$ respectively.  Given an energy $E$ and luminosity ${\cal L}$, $\Phi(t)$ and $x(t)$, which determines the two time-domain GW polarizations in Eqs.~\eqref{eq:hoftp} and~\eqref{eq:hoftc}, and $\Psi(f)$, which determines the two frequency-domain GW polarizations in Eqs.~\eqref{eq:hoffp} and~\eqref{eq:hoffc}, can be found from the following ordinary differential equations:
%\begin{eqnarray}
%\label{eq:dPhidt}
%\frac{d\Phi(t)}{dt}&=&\frac{x^{3/2}}{M},\\
%\label{eq:dxdt}
%\frac{d x(t)}{dt}&=&\frac{-{\cal L}}{E^{\prime}},
%\end{eqnarray}
%where $E^{\prime}=dE/dx$ is the analytic derivative of the energy $E$, and the luminosity ${\cal L}$ is related to energy via the energy balance requirement $dE/dt=-{\cal L}$.

%The energy and luminosity of two inspiralling point-particles can be easily solved for in the Newtonian limit of General Relativity.  However, as the system approaches merger, the orbital speed becomes relativistic, and this quasi-Newtonian limit is not enough to accurately describe its waveform.  We therefore use post-Newtonian (PN) theory, which involves expansions of the equations of motion in orders of orbital speed, to push beyond this limit.  Furthermore, compact binaries containing a neutron star noticeably depart from the point-particle approximation at high frequencies and require an additional correction to the energy and luminosity of the system relative to the point-particle terms.  More specifically, tidal interactions contribute to the phase of the orbit by adding a monopole-quadrupole interaction term to the energy.  This additional quadrupole moment adds to the total quadrupole moment of the system, which also increases the luminosity.
